---
title: "The wheat example from section 7.6 of the asreml manual (Butler et al. 2018)"
author: "Chris Brien"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document: default
---

```{r setup, include=TRUE}
library(asreml, quietly=TRUE)
library(asremlPlus)
suppressMessages(library(QTLRel, quietly=TRUE))
options(width = 100)
```

## Get data available in asremlPlus

```{r data, include = TRUE}
data(Wheat.dat)
```

## Fit the  initial model

In the following a model is fitted that has the terms that would be included for a balanced lattice. In addition, a term WithinColPairs has been included to allow for extraneous variation arising between pairs of adjacent lanes. Also, separable ar1 residual autocorrelation has been included. This model represents the maximal anticipated model, 

```{r first, include = TRUE}
current.asr <- asreml(yield ~ WithinColPairs + Variety, 
                      random = ~ Rep/(Row + Column) + units,
                      residual = ~ ar1(Row):ar1(Column), 
                      data=Wheat.dat)
```

The warning from `asreml` is probably dies to a bound term.

## Intialize a testing sequence by loading the current fit into an asrtests object

```{r initial, include = TRUE}
current.asrt <- as.asrtests(current.asr, NULL, NULL)

```

## Check for and remove any boundary terms

```{r check}
current.asrt <- rmboundary(current.asrt)
summary(current.asrt$asreml.obj)$varcomp
print(current.asrt, which = "testsummary")
```

Rep has been removed because it has been constrained to zero. Following the recommendation of Littel et al. (2006, p. 150), the bound on all variance components is set to unconstrained (U) so as to avoid bias in the estimate of the residual variance.

## Unbind Rep, Row and Column components and reload into an asrtests object

```{r unbind}
current.asr <- setvarianceterms(current.asr$call, 
                                terms = c("Rep", "Rep:Row", "Rep:Column"), 
                                bounds = "U")
current.asrt <- as.asrtests(current.asr, NULL, NULL)
current.asrt <- rmboundary(current.asrt)
summary(current.asrt$asreml.obj)$varcomp
print(current.asrt, which = "testsummary")
print(current.asrt, which = "pseudoanova")
```

Now the Rep component estimate is negative.

The `test.summary` output shows that no changes have been made to the model loaded using `as.asrtests`. The pseudo-anova table shows that Vaieties are highly significant ($p < 0.001$)

## Check term for within Column pairs (a post hoc covariate)

```{r covariate}
current.asrt <- testranfix(current.asrt, term = "WithinColPairs", 
                           drop.fix.ns=TRUE)
```

Generally, to determine what has been tested between two fits using `asreml` involves comparing two `asreml` calls and deciding what is different. For example what is the difference between the `asreml` call to fit the initial model and the following call?


`current.asr <- asreml(yield ~ Variety,`

                      `random = ~ Rep/(Row + Column) + units,
                      
                      residual = ~ ar1(Row):ar1(Column),
                      
                      data=Wheat.dat)`

On the other hand, it is clear from the `testranfix` call that the term `withinColPAirs` is being tested.

## Test nugget term

The nugget term represents non-spatial variance, such as measurement error. It is fitted using the `asreml` reserved word `units`.

```{r nugget}
current.asrt <- testranfix(current.asrt, "units", positive=TRUE)
```

## Test Row autocorrelation

We begin testing the autocorrelation by dropping the Row autocorrelation.

```{r rowar1}
current.asrt <- testresidual(current.asrt, "~ Row:ar1(Column)", 
                             label="Row autocorrelation", 
                             simpler=TRUE)
```

## Test Column autocorrelation (depends on whether Row autocorrelation retained)

The function `getTestPvalue` is used to get the p-value for the Row autocorrelation test. If it is significant then the Column autocorrelation is tested by by dropping the Column autocorrelation, while retaining the Row autocorrelation. Otherwise the model with just Row autocorrelation, whose fit is returned via `current.asrt` after the test, is compared to one with no autocorrelation. 

```{r colar1}
(p <- getTestPvalue(current.asrt, label = "Row autocorrelation"))
{ if (p <= 0.05)
  current.asrt <- testresidual(current.asrt, "~ ar1(Row):Column", 
                               label="Col autocorrelation", 
                               simpler=TRUE, update=FALSE)
  else
    current.asrt <- testresidual(current.asrt, "~ Row:Column", 
                                 label="Col autocorrelation", 
                                 simpler=TRUE, update=FALSE)
}
```

## Output the results

```{r results}
print(current.asrt, which = "test")
info <- infoCriteria(current.asrt$asreml.obj)
summary(current.asrt$asreml.obj)$varcomp
```

The `test.summary` shows is that the model with Row and without Column autocorrelation failed to converge. It is also clear that no changes were made to the variance terms. 

The `asreml.obj` in `current.asrt` contains the model selected by the testing process.

## Get current fitted asreml object and update to include standardized residuals

```{r update}
current.asr <- current.asrt$asreml.obj
current.asr <- update(current.asr, aom=TRUE)
Wheat.dat$res <- residuals(current.asr, type = "stdCond")
Wheat.dat$fit <- fitted(current.asr)
```

## Do diagnostic checking

### Do residuals-versus-fitted values plot

```{r resfit}
with(Wheat.dat, plot(fit, res))
```

### Plot variofaces

```{r VarioF}
variofaces(current.asr, V=NULL, units="addtores", 
           maxiter=50, update = FALSE)
```

The variofaces are the lag 1 plots of the sample semivariogram with simulated confidence envelopes (Stefanova et al., 2009). 

### Plot normal quantile plot

The plot is obtained using the `qqPlot` function from the  `QTLRel` function (Cheng, 2018).

```{r Nplot}
with(Wheat.dat, qqPlot(y = res, 
                       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles", 
                       main="Normal probability plot"))
```

## Get Variety predictions and all pairwise prediction differences and p-values

```{r WheatPreds}
Var.diffs <- predictPlus(classify = "Variety", 
                         asreml.obj=current.asr, 
                         error.intervals="halfLeast",
                         wald.tab=current.asrt$wald.tab, 
                         sortFactor = "Variety",
                         tables = "predictions")
```

We have set `error.intervals` to `halfLeast` so that the limits for $\pm o.5 LSD$ are calculated. When these are plotted overlapping error bars indicate predictions that are not significant, while those that do not overlap are significantly different (Snee, 1981).

Also set was `sortFactor`, so that the results would be ordered for the values of the predictions for Variety.

The function `predictPlus` returns an `alldiffs` object, a list consisting of the following components:

* `predictions`: the predictions, their standard errors and error intervals;

* `vcov`: the variance matrix of the predictions;

* `differences`: all pairwise differences between the predictions, 

* `p.differences`: p-values for all pairwise differences between the predictions;

* `sed`: the standard errors of all pairwise differences between the predictions; 

* `LSD`: the mean, minimum and maximum LSDs.


## Plot the Variety predictions, with halfLSD intervals, and the p-values

```{r plotPreds}
plotPredictions(Var.diffs$predictions, 
                classify = "Variety", y = "predicted.value", 
                error.intervals = "half")
plotPvalues(Var.diffs)
```

## References

Butler, D. G., Cullis, B. R., Gilmour, A. R., Gogel, B. J. and 
            Thompson, R. (2018). *ASReml-R Reference Manual Version 4.* VSN International Ltd, <http://asreml.org>.

Cheng, R. (2018) QTLRel: tools for mapping of quantitative traits of genetically related individuals and calculating identity coefficients from pedigrees. <https://CRAN.R-project.org/package=QTLRel>.

Gilmour, A. R., Thompson, R., & Cullis, B. R. (1995). Average Information REML: 
            An Efficient Algorithm for Variance Parameter Estimation in Linear 
            Mixed Models. *Biometrics*, **51**, 1440--1450. 

Snee, R. D. (1981). Graphical Display and Assessment of Means. *Biometrics*, **37**, 835--836. 

Stefanova, K. T., Smith, A. B. & Cullis, B. R. (2009) Enhanced diagnostics for the 
            spatial analysis of field trials. *Journal of Agricultural, Biological, 
            and Environmental Statistics*, **14**, 392--410.